Reagents: expressing and composing fine-grained concurrency

### 2011-10-17 paper/paper.tex
progress on background and reagents sections

\newcommand{\secref}[1]{\S\ref{sec:#1}}
\newcommand{\figref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\elide}[1]{}

\input{packages}

@@

To make this discussion concrete, we give a classic example of a fine-grained
concurrent data structure: Treiber's lock-free stack~\cite{?}, shown in
\figref{classic-treiber}.
%notice: changed to less desirable representation to make CAS work

@@

%% For example, containers like
%% stacks and queues provide communication; contructs like locks and
%% barriers provide synchronization; and (synchronous) message channels
%% provide both.

Treiber's stack scales better than a lock-based stack mainly because it
decreases the amount of shared data that must be updated per operation:
instead of acquiring a shared lock, altering a shared stack pointer, and
releasing the shared lock, Treiber's stack does a single CAS to update the
shared pointer directly.  Scalability can be further improved by

@@

First we give an overview of the library, and then we illustrate its
expressiveness and composability by showing how to implement several

@@

@@

\begin{itemize}
\item TreiberStack
\item with blocking
\item with elimination backoff
\item MSQueue
\item with blocking
\item with buckets?
\item Counter to semaphore
\item RWLocks
\item Tree barriers
\item Dining philosophers(ish)
\item Flat combining
\end{itemize}

@@

\newcommand{\ra}{\rightarrow}
\newcommand{\gives}{\vdash}
\newcommand{\proves}{\vdash}
\newcommand{\kw}[1]{\textsf{#1}}
\newcommand{\GA}{\ |\ }
\newcommand{\step}{\longrightarrow}
\newcommand{\cstep}[1]{\stackrel{#1}{\Longrightarrow}}

\newcommand{\evalCtx}{\mathbb{E}}
\newcommand{\procCtx}{\mathbb{P}}

\newcommand{\ret}[1]{\kw{ret}(#1)}
\newcommand{\never}{\kw{never}}
\newcommand{\postCommit}[1]{\kw{postCommit}(#1)}
\newcommand{\computed}[1]{\kw{computed}(#1)}
\newcommand{\compose}{{\small\texttt{>=>}}}
\newcommand{\lift}[1]{\kw{lift}(#1)}
\newcommand{\send}[1]{#1}
\newcommand{\cas}[3]{\kw{cas}(#1,#2,#3)}
\newcommand{\rd}[1]{\kw{read}(#1)}
\newcommand{\newChan}{\kw{new Chan}}
\newcommand{\newRef}[1]{\kw{new Ref(#1)}}

@@

>>>>

\input{packages}
\input{macros}

@@

\figref{classic-treiber} gives a classic example of a fine-grained
concurrent data structure: Treiber's lock-free stack~\cite{?}.  The
stack is represented as an \emph{immutable} linked list.  Mutation
occurs solely through the \lstinline{head} pointer, represented here
as an \lstinline{AtomicReference}.  The \lstinline{head} is mutated
using \lstinline{compareAndSet} (here abbreviated as \lstinline{cas}),
which takes an expected value and a new value, and atomically updates
the reference

%notice: changed to less desirable representation to make CAS work

@@

Treiber's stack scales better than a lock-based stack mainly because
it decreases the amount of shared data that must be updated per
operation: instead of acquiring a shared lock, altering a shared stack
pointer, and releasing the shared lock, Treiber's stack does a single
CAS to update the shared pointer directly.  But the head pointer of
the stack is still a centralized source of contention.  Spreading out
the contention requires parallelizing stack operations---but how,
given that they all involve modifying the top of the stack?

Parallelization requires a change of perspective.  Normally, we view
concurrent operations on the same part of a data structure as
competing to atomically update that data structure; we want the
operations to be isolated.  But if

@@

In the last section, we saw a range of strategies for communication
and synchronizion at a fine grain:
\begin{itemize}
\item using CAS to make direct, atomic updates;
\item using exponential backoff to ease contention; 
\item using side-channel communication to increase parallelism; 
\item using disjoint spinwaiting to eliminate useless cache traffic; 
\item using dual data structures to incorporate blocking.
\end{itemize}
Reagents unify these disparate techniques into a single framework for
writing concurrent algorithms.  Some of the techniques are built into
the framework, while others are expressible as reusable abstractions.
% extensibility and composability

\subsection{The main idea}

%% The aim of reagents is to make it easy to express these and other
%% patterns, while providing compositionality blah blah

%% First we give an overview of the library, and then we illustrate its
%% expressiveness and composability by showing how to implement several

@@

\subsection{Expressiveness by example}

@@

\begin{itemize}
\item TreiberStack
\item with blocking
\item with elimination backoff
\item MSQueue
\item with blocking
\item with buckets?
\item Counter to semaphore
\item RWLocks
\item Tree barriers
\item Dining philosophers(ish)
\item Flat combining - doesn't work if commit is two-phase process.
  probably too much to bring into scope for this paper, anyway
\end{itemize}

\begin{lstlisting}
class MSQueue[A] {
  private final case class Node(
    data: A, next: Ref[Node] = Ref(null)
  )
  private val head = Ref(Node(null))
  private val tail = Ref(head.read!())

  val enq: Reagent[A, Unit] = computed { (x:A) =>
    @tailrec def search: Reagent[Unit,Unit] = 
      tail.read ! () match {
        case    Node(_, r@Ref(null)) => 
          r.cas(null, Node(x))
        case ov@Node(_, Ref(nv)) => 
          tail.cas(ov,nv) !? (); search
      }
    search
  }
  val tryDeq: Reagent[Unit, Option[A]] = head.upd {
    case Node(_, Ref(n@Node(x, _))) => (n, Some(x))
    case emp => (emp, None)
  }
}
\end{lstlisting}

@@

Try to keep this minimal for the sake of space

Purpose of the semantics is mainly to clarify the relationship between
the shared-state and message-passing aspects of reagents.

\begin{itemize}
  \item read, cas
  \item send
  \item choice
  \item react
  \item bind and ret
  \item computed?
\end{itemize}

@@

  - somewhere say that: we examine fairly simple examples in the
  paper, due to size constraints.  but we can handle more complex
  cases like sets/maps represented via skip lists, basket queues, 
  and are exploring flat combining.
